---
title: "Effects without a Cause: The Search for Demographic Anomalies"

# to produce blinded version set to 1
blinded: 0

authors: 
- name: Mathew E. Hauer
  thanks:  Thanks y'all!
  affiliation: Department of Sociology, Florida State University
  
- name: Stephanie A. Bohon
  affiliation: Department of Sociology, University of Tennessee - Knoxville

keywords:
- mortality
- fertility
- causal inference

abstract: |
  The proliferation of data, modern computing advances, and powerful statistical algorithms unlock the potential search for hidden, previously un- or understudied demographic anomalies. Here, we demonstrate how demographers can use causal inference techniques to identify anamolies by investigating US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?. We find 38 states exhibited at least one mortality anomaly, totalling more than 318k anomalous deaths and an additional 6 states exhibited at least one mortality non-plague? totalling 164k protective deaths. 12 states exhibited baby busts, totalling more than 240k missing births and an additional 11 states exhibited baby booms totalling more than 134k additional births. These results suggest the widespread detection of demographic anomalies. Our analysis does not examine the *causes* of these anomalies and our results point to important further research on the causes of anomalous demographic behavior.    
  
bibliography: mybibfile
output: rticles::asa_article
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}

---

\newpage

```{r setup, include=FALSE}
rm( list=ls() )
knitr::opts_chunk$set(echo  = TRUE)

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
read_chunk('../R/SCRIPTS/001-DataLoad.R')
read_chunk('../R/SCRIPTS/002-StatesAnalysis.R')
read_chunk('../R/SCRIPTS/003-SummaryStats.R')
```

```{r libraries, include=FALSE}
```

```{r DataLoad, include=FALSE}
```

```{r SumStats, include=FALSE}
```

# Introduction

The proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms mark the era of "Big Data" or data science [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography]. While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena. However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden].

Causal inference, in its most common usage, is an attempt to uncover the underlying mechanism that results in changes in a phenomenon and have a long history approaches in social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment. For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph]. Using natural experiments provide less certainty about causation and fewer opportunities to replicate work but they provide important and widespread insights such as the determinants of migration, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use]. Both RCTs and natural experiments follow traditional, hypothesis testing, inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.

Both RCTs and natural experiments follow traditional hypothesis testing--inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed. Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations
[@pearl2018book].

Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017]. In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big]. Owing to the long history of big data sets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause. In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].

In this paper, we use modern statistical outlier detection algorithms [@chen1993joint] on nearly twenty years of mortality and fertility data at the US state-level to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask two questions regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, demographers may be able to better forecast populations and provide policy solutions for impending problems.

# Materials and Methods

## Method
We use a statistical time series outlier detection algorithm [@chen1993joint], implemented in the R programming language [@rcore] via the tsoutliers package [@tsoutliers2019]. This algorithm iteratively uses ARIMA models to 1) produce a counter-factual time series to initially detect an outlier or anomaly, and 2) refit the ARIMA with the outliers removed. Here we briefly summarize and describe the method.

Often, the behavior of a time series can be described and summarized in ARIMA models. If a series of values, $y_t^*$, is subject to $m$ interventions or outliers at time points $t_1,t_2,…,t_m$, then $y_t^*$ can be defined as

$$y_t^* = \sum_{j=1}^{m} \omega_jL_j(B)I_t(t_j) + \frac{\theta(B)}{\phi(B)\alpha(B)}\alpha_t$$

Where
$I_t(t_j)$ is an indicator variable with a value of 1 at observation $t_j$ and where the $j$th outlier arises,
$\phi(B)$ is an autoregressive polynomial with all roots outside the unit circle,
$\theta(B)$ is a moving average polynomial with all roots outside the unit circle,
and $\alpha(B)$ is an autoregressive polynomial with all roots on the unit circle.

We examine three types of outliers at time point $t_m$:
1) additive outliers (AO), defined as $L_j(B)=1$;
2) level shift outliers (LS), defined as $L_j(B) = 1/(1-B)$; and
2) temporary change outliers (TC), defined as $L_j(B) = 1/(1-\delta B)$.

Colloquially, additive outliers arise when a single event causes the time series to unexpectedly increase/decrease for a single time period; level shift outliers arise when an event causes the time series to unexpectedly increase/decrease for multiple time periods; and temporary change outliers arise when an event causes the time series to unexpectedly increase/decrease with lingering effects that decay over multiple time periods.

An outlier is detected using a regression equation

$$ \pi(B)y_t^* \equiv \hat{e} = \sum_{j=1}^m \omega_j \pi(B)L_j(B)I_t(t_j) + \alpha_t $$

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers. 

We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series without outliers removed, from $y$, the original, unadjusted time series.

## Data

We search for demographic anomalies using the Center for Disease Control and Prevention’s online WONDER monthly fertility (2003-2017) and mortality databases (1999-2016) for all fifty states and the District of Columbia
[@CDC_fert07;@CDC_mort]. These data sets contain every birth and death record in the United States over the time periods of interest, representing the universe of both mortality and fertility data in the US.  These data are considered the “gold standard” of data collections [@mahapatra2007civil] and have been considered “complete” since 1968 [@hetzel2016us]. We search over each state equivalent’s (n=51) mortality (n=228) and fertility (n=180) monthly time series for a total of 20,808 state-months of data.

# Results
We detect numerous anomalous mortality and fertility events at the US state-level since 1999. A full listing of these anomalies can be found in the **Supplementary Materials**. We begin with a summary of the anomalies we detect and then we highlight highly significant mortality and fertility anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change Outliers) with plausible explanations. Finally, we highlight two strong anomalies that bely explanation.

## Overall Anomalies


```{r, echo = FALSE}

sums2 <- sums %>%
  dplyr::select(percent_anom = `\\% of Total`, everything())
kable(sums,
       "latex",
       booktabs = T,
       format.args = list(big.mark = ","),
       escape = FALSE,
       caption = "\\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component. \\label{sumtable}") %>%
   kable_styling(position = "center")

```

**\autoref{sumtable}** report the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types. We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`). Given that anomalies are always related to events [@Song2018ACM], the finding of more mortality than fertility anomalies is not surprising.  Mortality is more easily subject to instantaneous change (like a natural disaster) compared with fertility, and fertility is linked to human decision-making more strongly than mortality [@Stein2014Fertility]. Previous studies document that some events, like the 2005 Oklahoma City bombings, resulted in both fertility and mortality changes [@Rogers2005OKBombing], but changes in fertility manifest approximately 9-months after an event, and the relationship between a fertility-inducing event and behavior change (such as a prolonged blizzard or power blackout) is less strong than the relationship between a mortality-inducing event (such as an earthquake) and death. For example, Evans and his colleagues [@Evans2008Hurricanebirth] show that catastrophic weather events can lead to increased fertility, but some events reduce fertility.  Furthermore, the effect is seen primarily in those who already have one child, with childless people unlikely to have a child due to a weather event.  

We also find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`). Fertility anomalies overwhelmingly tend toward lower fertility, with only a few fertility anomalies yielding more births. This is consistent with research findings that link high severity weather events to significantly lower fertility [@Evans2008Hurricanebirth]. Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than less deaths. Again, this is not surprising, as the kinds of events that increase death (e.g., plant closings are associated with increased opioid death [@Venkataramani2019]) are more common than the events that decrease death (e.g., cannabis legalization is associated with decreased opioid deaths [@Livingston2017Cannabis]) These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.

**\autoref{fig:summap}** maps these results. Three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota. Another nine states exhibited just a single anomaly (DE, DC, ID, IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts exhibited the most anomalies with nine.



```{r AnomalyMap, echo= FALSE, message = FALSE, warning = FALSE, fig.height=7, fig.cap= paste("\\textbf{Summary of Anomalies by type, demographic component, and State.} We observe much fewer fertility anomalies than mortality anomalies. \\label{fig:summap}")}

data(state_laea)
anom_mort <- read_rds("../R/DATA-PROCESSED/mortality_anomalies") %>%
  na.omit %>%
  mutate(component = "Mortality")

anom_fert <- read_rds("../R/DATA-PROCESSED/fertility_anomalies") %>%
  na.omit %>%
  mutate(component = "Fertility")

anom_mort2 <- rbind(anom_mort, anom_fert) %>%
  group_by(type, location, component) %>%
  dplyr::summarise(Anomalies =  n()) %>%
  mutate(ident = paste0(component, " - ", type),
         GEOID = substr(location, 2,3)) %>%
  ungroup() %>%
  dplyr::select(GEOID, Anomalies, ident) %>%
  spread(ident, Anomalies) %>%
    mutate_all(funs(replace_na(.,0)))


states <- left_join(state_laea, anom_mort2) %>%
  gather(ident, Anomalies, `Fertility - AO`:`Mortality - TC`) %>%
  mutate_all(funs(replace_na(.,0))) %>%
  mutate(Anomalies = factor(Anomalies))

states$ident2 <- factor(
  states$ident,
  levels = c("Mortality - AO", "Fertility - AO", 
             "Mortality - TC", "Fertility - TC", 
             "Mortality - LS", "Fertility - LS")
)

colorpal <- c("#ffffff", "#ffffcc", "#a1dab4", "#41b6c4", "#2c7fb8", "#253494")

  ggplot(data=states) +
  geom_sf(aes(fill = Anomalies),  lwd = 1/2) +
  theme_bw() +
  theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        legend.position="bottom"
  )+
  guides(fill=guide_legend(nrow=1,byrow=TRUE)) +
  facet_wrap(~ident2, ncol=2) +
    scale_fill_manual(name = "# of Anomalies", values = colorpal) +
    
  NULL
```

## Mortality 
We begin with mortality for New York State (**\autoref{fig:mortnewhamp}a**). We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies highly significant. The algorithm correctly identifies September 2001 as an additive outlier (2001:09 t=6.40) where mortality in that month was 1,628 higher than anticipated. This mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies. 

In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 t=5.59) which prevented almost 870 deaths per month. This shift totals more than 144,000 averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states. This translates to 4.5% fewer deaths than expected over the time period. What is driving this mortality protection? What policies did NY put into place that might have contributed to this considerable mortality reduction? What environmental conditions may have changed? These are the kinds of questions that arise from our analyses.

Contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}b**). Here we detect two highly significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68). These anomalies suggest New Hampshire experienced 9,700 more deaths (+14% more than expected) in a seven-year period beginning in early 2010. This is the single largest percentage mortality increase/decrease we detected among all states. Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is reasonable that we detect this epidemic in our results.

```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 5, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies. The bottom part of each panel contains the magnitude and type of the outlier in red. In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004. In New Hampshire (b), we detect two outliers, both level shift outliers (LS) in April 2010 and again in November 2014. These anomalies suggest New York experienced a significant mortality event in September 2001 and New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years. \\label{fig:mortnewhamp}")}

a <-  function() {
plot(outlier.mort.ny, cex = 0.8)
title("New York - Mortality", adj = 0.9, line =2)
}

b <-  function() {
  plot(outlier.mort.nh, cex = 0.8)
  title("New Hampshire - Mortality", adj = 0.82,line =2)
}

plot_grid(ggdraw(a) ,
               ggdraw(b), 
               ncol=1,
               labels= "auto")
```



## Fertility

We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**). Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[1], digits = 3)`; and 2005:10 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[2], digits = 3)`), coinciding with the destruction of Hurricane Katrina from the same year. The migration associated with Hurricane Katrina has garnered most of the attention of demographers [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana. We estimate `r prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.

We also highlight a level shift (LS) outlier for fertility in Connecticut. Here we detect a strong (*t*= `r prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)`; `r prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertla}b**). This LS toward lower fertility reduced the # of births by `r prettyNum(sum(outlier.fert.conn$y - outlier.fert.conn$yadj), digits = 0, big.mark = ",")` or `r percent(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180]))` lower than the counterfactual time series. The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occured just 11 months before we detect a shift toward lower fertility. We believe the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked.

```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane. In (b), we detect one outlier, a level shift outlier (LS) in August 2009. We believe this reduction is attributable to the stock market crash 11 months earlier. \\label{fig:fertla}")}

a <-  function() {
  plot(outlier.fert.la, cex.axis=0.8)
  title(main = "Louisiana - Fertility", adj = 0.9)
}

b <- function(){
  plot(outlier.fert.conn, cex.axis=0.8) +
  title(main = "Connecticut - Fertility", adj = 0.88)
}

plot_grid(ggdraw(a) ,
               ggdraw(b), 
               ncol=1,
               labels= "auto")
```

## Interesting Anomalous Fertility/Mortality events
In the examples above, we highlighted four fertility/mortality anomalies with plausible explanations. In the case of New York and New Hampshire, the mortality anomalies have plausible explanations. It seems likely that New York’s AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise in New Hampshire’s mortality starting in 2010 could be linked to the opioid epidemic. Similarly, Louisiana's TC anomalies seem linked to Hurricanes Katrina and Rita while the LS anomaly in Connecticut's fertility appears linked to the Great Recession. However, we detect numerous seemingly unexplainable demographic anomalies in other states. **\autoref{fig:ferthawaii}** shows two such unexplained anomalies. 

In **\autoref{fig:ferthawaii}a** we identify a single additive anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with a *t*-statistic of `r prettyNum(outlier.fert.hi$outliers$tstat, digits = 3)`, 14% above the counter-factual time series. This single, anomalous month is also the second highest monthly births in the time series. We have no plausible explanation for this anomaly. We do not believe this is simply a data error as the other extreme values, September 2008 and February 2005 with the highest and lowest recorded fertility respectively, were not identified as anomalous events. Even if we were to assume this anomaly resulted from data entry error, it remains an *unaltered* data error in the Hawaiian monthly fertility data.

This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r prettyNum(outlier.mort.oh$outliers$tstat[7], digits = 3)`). This LS is more than 1,100 deaths per month less than the counterfactual time series, suggesting had nearly 40,000 fewer deaths since February 2015 than expected. This is the single-largest LS among all states. We could not identify the potential policies Ohio might have put into place to provide such a strong mortality protection.

```{r TrueAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaii state fertility (a) and Ohio state mortality (b).} Here we detect one outlier, an additive outlier (AO) in May 2014 in Hawaii (a), representing a large, unexplainable 14% increase in expected births in that month. We also detect a strong level shift (LS) reduction (b), representing a large, unexplainable reduction of nearly 40,000 deaths in Ohio.  \\label{fig:ferthawaii}")}

a <-  function() {
plot(outlier.fert.hi, cex.axis=0.8)
  title(main = "Hawaii - Fertility", adj = 0.88)
}

b <-  function() {
plot(outlier.mort.oh, cex.axis=0.8)
  title(main = "Ohio - Mortality", adj = 0.9)
}


plot_grid(ggdraw(a) ,
               ggdraw(b), 
               ncol=1,
               labels= "auto")
```




\newpage
